---
title: "Practical Machine Learning Course Project"
author: "Jose Augusto Barros de Oliveira"
date: "02/01/2021"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (skimr)
library(caret)
library(lubridate)
library(corrplot)
```

## Summary  

The goal of this project is to predict whether weight-lift exercises were performed as recommended. It uses the experimental design and data from this project on [Human Activity Recognition] (http://groupware.les.inf.puc-rio.br/har).
The dataset contains 160 features and 19622 cases. After some Exploratory Data Analysis (EDA), we decide to go through modeling with 60 features, besides the target. We performed Principal Component Analysis (PCA) in order to reduce the dimensionality, ending with 20 components that explained 90% of the variance.
The data was splited in two sets, the training one with 70% of cases.
We choose a Xg Boost model and used cross validation (CV) with K = 10 to tuning parameters.
The overall model accuracy on test set was xx% and the Confusion Matrix showed better results than the original paper.


## Data ingestion and first analysis

```{r dataingestion}
# Train
theURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

if(file_test("-f", "pml-training.csv") == FALSE) 
  {
  download.file(theURL,"pml-training.csv",mode = "wb")
}

traindata <- read.csv("pml-training.csv", na.strings = c("","NA","NULL"))

# Test
theURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if(file_test("-f", "pml-testing.csv") == FALSE) 
{
  download.file(theURL,"pml-testing.csv",mode = "wb")
}

testdata <- read.csv("pml-testing.csv", na.strings = c("","NA","NULL"))

# Data assessment
dataquality <- skim_without_charts(traindata)
dataquality

```
Mostly features are empty or near that so we kept the ones carrying some info.
We will use non parametric methods, so the assumption of normality of the features doesn't matter. But the inspection of qqplots shows that mostly are gaussian with heavy tails


